---
title: "TCR Diversity Validation Report"
subtitle: "NanoString-style TRBV Proxy vs True Clonotype Diversity — Blum, Lechner & Luoma Datasets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    self_contained: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)
```

```{r libs, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(ggrepel)
  library(scales)
  library(data.table)
  library(RColorBrewer)
  library(patchwork)
  library(pheatmap)
  library(plotly)
  library(DT)
  library(knitr)
  library(kableExtra)
})

# Source functions from repo root
REPO_ROOT <- here::here()
source(file.path(REPO_ROOT, "R", "tcr_diversity_functions.R"))

VAL_DIR  <- file.path(REPO_ROOT, "validation")
MIN_CELLS    <- 30
DOWNSAMPLE_N <- 200

DATASET_COLORS <- c(
  "blum"    = "#2166AC",
  "lechner" = "#D6604D",
  "luoma"   = "#4DAF4A"
)
```

```{r load-helpers, include=FALSE}
# ── h5ad loading ──────────────────────────────────────────────────────────────
load_h5ad_metadata <- function(path) {
  if (requireNamespace("zellkonverter", quietly = TRUE)) {
    sce  <- zellkonverter::readH5AD(path, X = FALSE, use_hdf5 = FALSE)
    meta <- as.data.frame(SingleCellExperiment::colData(sce))
    return(tibble::rownames_to_column(meta, "barcode_aggr"))
  }
  if (requireNamespace("anndata", quietly = TRUE)) {
    ad   <- anndata::read_h5ad(path)
    meta <- as.data.frame(ad$obs)
    return(tibble::rownames_to_column(meta, "barcode_aggr"))
  }
  stop("Install zellkonverter (BiocManager) or anndata (CRAN) to load .h5ad files.")
}

`%||%` <- function(a, b) if (!is.null(a)) a else b

.TRBV_PATTERNS  <- c("^v_gene$","^trbv$","^trbv_gene$","^v_gene_beta$",
                      "IR_VDJ_1_v_call","^TCR_Beta_V_gene$","^beta_v_gene$",
                      "^vb_gene$","^TRBV_gene$","v_call")
.CLONO_PATTERNS <- c("^clonotype_id$","^clone_id$","^cloneid$",
                      "^cdr3_beta$","^cdr3_b_aa$","^IR_obs_vj_1_junction_aa$",
                      "^TCR_clonotype$","^clonotype$","^clone$",
                      "^raw_clonotype_id$","^ct_id$")
.SAMPLE_PATTERNS <- c("^sample_id$","^sample$","^donor$","^patient$",
                       "^patient_id$","^donor_id$","^orig.ident$",
                       "^batch$","^library_id$")

.find_col <- function(nms, pats) {
  for (p in pats) { h <- grep(p, nms, ignore.case=TRUE, value=TRUE); if (length(h)) return(h[1]) }
  NULL
}

normalise_tcr_metadata <- function(meta, dataset_id,
                                    trbv_col=NULL, clono_col=NULL, sample_col=NULL) {
  cols <- colnames(meta)
  trbv_col   <- trbv_col   %||% .find_col(cols, .TRBV_PATTERNS)
  clono_col  <- clono_col  %||% .find_col(cols, .CLONO_PATTERNS)
  sample_col <- sample_col %||% .find_col(cols, .SAMPLE_PATTERNS)

  if (!is.null(trbv_col)   && trbv_col   != "trbv_gene")
    meta <- meta %>% rename(trbv_gene    = all_of(trbv_col))
  if (!is.null(clono_col)  && clono_col  != "clonotype_id")
    meta <- meta %>% rename(clonotype_id = all_of(clono_col))
  if (!is.null(sample_col) && sample_col != "sample_id")
    meta <- meta %>% rename(sample_id    = all_of(sample_col))
  else if (is.null(sample_col))
    meta$sample_id <- dataset_id

  if (!"trbv_gene"    %in% colnames(meta)) meta$trbv_gene    <- NA_character_
  if (!"clonotype_id" %in% colnames(meta)) meta$clonotype_id <- NA_character_
  meta$dataset <- dataset_id

  attr(meta, "col_map") <- list(trbv=trbv_col, clono=clono_col, sample=sample_col)
  meta
}

load_and_prep <- function(fname, dataset_id, ...) {
  path <- file.path(VAL_DIR, fname)
  raw  <- load_h5ad_metadata(path)
  normalise_tcr_metadata(raw, dataset_id, ...)
}
```

```{r load-data, cache=TRUE, include=FALSE}
blum_meta    <- load_and_prep("blum_combined_tissue_data.h5ad",                      "blum")
lechner_meta <- load_and_prep("lechner_thyroiditis_tissue_cd8_seurat_object.h5ad",   "lechner")
luoma_meta   <- load_and_prep("luoma_colitis_cd8_seurat_object.h5ad",                "luoma")

all_datasets <- list(blum = blum_meta, lechner = lechner_meta, luoma = luoma_meta)
```

```{r run-analysis, cache=TRUE, include=FALSE}
run_ds <- function(meta, gv = "sample_id") {
  run_tcr_diversity_analysis(meta, group_vars=gv,
                              min_cells=MIN_CELLS, downsample_n=DOWNSAMPLE_N)
}

blum_res    <- run_ds(blum_meta)
lechner_res <- run_ds(lechner_meta)
luoma_res   <- run_ds(luoma_meta)

results_list <- list(blum = blum_res, lechner = lechner_res, luoma = luoma_res)
```

---

## Overview

This report validates the **NanoString-style TRBV usage diversity proxy** against true CDR3-level clonotype diversity across three published single-cell TCR sequencing datasets.

> **Important:** The TRBV-based metric measures V-gene usage breadth/evenness. It **cannot** resolve CDR3 clonotypes, antigen specificity, or clonal expansion within a TRBV family. It is labelled throughout as *TRBV usage diversity proxy*, not *true TCR diversity*.

| Dataset | Description | Tissue | T cell subset |
|---------|-------------|--------|----------------|
| **Blum** | Blum et al. — combined tissue | Multiple | Mixed |
| **Lechner** | Lechner et al. — thyroiditis | Thyroid | CD8 |
| **Luoma** | Luoma et al. — colitis | Gut | CD8 |

---

## 1  Dataset Summaries {.tabset}

```{r coverage-fn, include=FALSE}
coverage_table <- function(meta, dataset_id) {
  n   <- nrow(meta)
  cm  <- attr(meta, "col_map")
  tibble(
    Metric  = c("Total cells", "Cells with TRBV gene", "Cells with clonotype ID",
                "Unique samples", "Unique TRBV genes", "Unique clonotypes",
                "Detected TRBV column", "Detected clonotype column", "Detected sample column"),
    Value   = c(
      scales::comma(n),
      paste0(scales::comma(sum(!is.na(meta$trbv_gene) & meta$trbv_gene != "")),
             sprintf(" (%.1f%%)", 100*mean(!is.na(meta$trbv_gene) & meta$trbv_gene != ""))),
      paste0(scales::comma(sum(!is.na(meta$clonotype_id) & meta$clonotype_id != "")),
             sprintf(" (%.1f%%)", 100*mean(!is.na(meta$clonotype_id) & meta$clonotype_id != ""))),
      as.character(n_distinct(meta$sample_id)),
      as.character(n_distinct(meta$trbv_gene, na.rm=TRUE)),
      as.character(n_distinct(meta$clonotype_id, na.rm=TRUE)),
      cm$trbv   %||% "not detected",
      cm$clono  %||% "not detected",
      cm$sample %||% paste0("none — used '", dataset_id, "'")
    )
  )
}
```

### Blum
```{r blum-coverage}
coverage_table(blum_meta, "blum") %>%
  kable(col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

### Lechner
```{r lechner-coverage}
coverage_table(lechner_meta, "lechner") %>%
  kable(col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

### Luoma
```{r luoma-coverage}
coverage_table(luoma_meta, "luoma") %>%
  kable(col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

---

## 2  Diversity Metric Tables {.tabset}

Columns: `H_TRBV` = Shannon diversity of TRBV gene usage; `H_clonotype` = Shannon diversity of CDR3 clonotypes; `true_clonality` = 1 − normalised Shannon (high = oligoclonal).

```{r summary-tables, include=FALSE}
make_summary <- function(res) build_summary_table(res$comparison)
blum_tbl    <- make_summary(blum_res)
lechner_tbl <- make_summary(lechner_res)
luoma_tbl   <- make_summary(luoma_res)
```

### Blum
```{r blum-tbl}
datatable(blum_tbl, rownames=FALSE, filter="top",
          options=list(pageLength=10, scrollX=TRUE),
          caption="Blum — per-sample TRBV proxy and true clonotype diversity") %>%
  formatRound(where(is.numeric, blum_tbl), digits=4)
```

### Lechner
```{r lechner-tbl}
datatable(lechner_tbl, rownames=FALSE, filter="top",
          options=list(pageLength=10, scrollX=TRUE),
          caption="Lechner — per-sample TRBV proxy and true clonotype diversity") %>%
  formatRound(where(is.numeric, lechner_tbl), digits=4)
```

### Luoma
```{r luoma-tbl}
datatable(luoma_tbl, rownames=FALSE, filter="top",
          options=list(pageLength=10, scrollX=TRUE),
          caption="Luoma — per-sample TRBV proxy and true clonotype diversity") %>%
  formatRound(where(is.numeric, luoma_tbl), digits=4)
```

---

## 3  TRBV Gene Usage {.tabset}

Stacked barplots of the top 20 TRBV genes per sample. Each bar = one sample; colours = TRBV gene families.

### Blum
```{r blum-bar, fig.height=6}
plot_trbv_stacked_bar(blum_meta, group_var="sample_id", top_n_genes=20)
```

### Lechner
```{r lechner-bar, fig.height=6}
plot_trbv_stacked_bar(lechner_meta, group_var="sample_id", top_n_genes=20)
```

### Luoma
```{r luoma-bar, fig.height=6}
plot_trbv_stacked_bar(luoma_meta, group_var="sample_id", top_n_genes=20)
```

---

## 4  TRBV Gene Usage Heatmaps {.tabset}

Proportion of cells per sample assigned to each of the top 25 TRBV genes. Rows and columns are hierarchically clustered.

### Blum
```{r blum-heat, fig.height=7}
plot_trbv_heatmap(blum_meta, group_var="sample_id", top_n_genes=25)
```

### Lechner
```{r lechner-heat, fig.height=7}
plot_trbv_heatmap(lechner_meta, group_var="sample_id", top_n_genes=25)
```

### Luoma
```{r luoma-heat, fig.height=7}
plot_trbv_heatmap(luoma_meta, group_var="sample_id", top_n_genes=25)
```

---

## 5  Clonotype Rank-Abundance Curves {.tabset}

Log–log rank-abundance (Zipf) plots. A steeper curve indicates a more clonal repertoire; a shallower curve indicates greater diversity.

### Blum
```{r blum-rank, fig.height=5}
plot_clonotype_rank_abundance(blum_meta, group_var="sample_id")
```

### Lechner
```{r lechner-rank, fig.height=5}
plot_clonotype_rank_abundance(lechner_meta, group_var="sample_id")
```

### Luoma
```{r luoma-rank, fig.height=5}
plot_clonotype_rank_abundance(luoma_meta, group_var="sample_id")
```

---

## 6  TRBV Proxy vs True Clonotype Diversity {.tabset}

Each point = one sample. The **red dotted line** is the 1:1 reference. The grey band is a linear regression ± SE. Spearman ρ is shown in the subtitle.

> A high ρ means the TRBV proxy is a reasonable surrogate in this context. A low ρ or points far from the 1:1 line indicate failure modes.

### Blum
```{r blum-scatter, fig.height=6}
plot_h_trbv_vs_h_clonotype(blum_res$comparison$table, group_var="sample_id")
```

### Lechner
```{r lechner-scatter, fig.height=6}
plot_h_trbv_vs_h_clonotype(lechner_res$comparison$table, group_var="sample_id")
```

### Luoma
```{r luoma-scatter, fig.height=6}
plot_h_trbv_vs_h_clonotype(luoma_res$comparison$table, group_var="sample_id")
```

### All datasets
```{r all-scatter, fig.height=5, fig.width=13}
all_comp <- map_dfr(results_list, ~ .x$comparison$table, .id="dataset")

sp_all <- tryCatch(cor.test(all_comp$H_TRBV, all_comp$H_clonotype,
                             method="spearman", exact=FALSE), error=function(e) NULL)
sub_txt <- if (!is.null(sp_all))
  paste0("Overall Spearman ρ = ", round(sp_all$estimate, 3),
         "  (p = ", signif(sp_all$p.value, 3),
         ", n = ", nrow(all_comp), " samples across 3 datasets)")
else ""

ggplot(all_comp, aes(x=H_TRBV, y=H_clonotype,
                      color=dataset, label=sample_id)) +
  geom_point(size=3, alpha=0.85) +
  geom_smooth(method="lm", se=TRUE, color="grey35",
              linetype="dashed", linewidth=0.7) +
  geom_text_repel(size=2.5, max.overlaps=15) +
  geom_abline(slope=1, intercept=0, color="red",
              linetype="dotted", linewidth=0.6) +
  facet_wrap(vars(dataset), scales="free") +
  scale_color_manual(values=DATASET_COLORS, name="Dataset") +
  labs(title   = "TRBV Usage Diversity vs True Clonotype Diversity — All Datasets",
       subtitle = sub_txt,
       x = expression(H[TRBV]~"(TRBV usage diversity proxy)"),
       y = expression(H[clonotype]~"(true CDR3-level diversity)"),
       caption = "Red dotted line: 1:1 reference. TRBV proxy cannot resolve CDR3 clonotypes.") +
  theme_bw(base_size=12) + theme(legend.position="bottom")
```

---

## 7  Clonality Proxy vs True Clonality {.tabset}

Clonality = 1 − normalised Shannon. A value near 1 means a highly oligoclonal repertoire; near 0 means highly diverse.

### Blum
```{r blum-clon, fig.height=6}
plot_clonality_comparison(blum_res$comparison$table, group_var="sample_id")
```

### Lechner
```{r lechner-clon, fig.height=6}
plot_clonality_comparison(lechner_res$comparison$table, group_var="sample_id")
```

### Luoma
```{r luoma-clon, fig.height=6}
plot_clonality_comparison(luoma_res$comparison$table, group_var="sample_id")
```

### All datasets
```{r all-clon, fig.height=5, fig.width=13}
ggplot(all_comp, aes(x=TRBV_clonality_proxy, y=true_clonality,
                      color=dataset, label=sample_id)) +
  geom_point(size=3, alpha=0.85) +
  geom_smooth(method="lm", se=TRUE, color="grey35",
              linetype="dashed", linewidth=0.7) +
  geom_text_repel(size=2.5, max.overlaps=15) +
  geom_abline(slope=1, intercept=0, color="red",
              linetype="dotted", linewidth=0.6) +
  facet_wrap(vars(dataset), scales="free") +
  scale_x_continuous(limits=c(0,1)) +
  scale_y_continuous(limits=c(0,1)) +
  scale_color_manual(values=DATASET_COLORS, name="Dataset") +
  labs(title = "TRBV Clonality Proxy vs True Clonality — All Datasets",
       x = "TRBV clonality proxy (1 − H_norm_TRBV)",
       y = "True clonality (1 − H_norm_clonotype)") +
  theme_bw(base_size=12) + theme(legend.position="bottom")
```

---

## 8  Bland-Altman Analysis {.tabset}

Bias (blue line) and ±1.96 SD limits of agreement (red dashed) between H_TRBV and H_clonotype. Points above zero mean the TRBV proxy overestimates diversity.

### Blum
```{r blum-ba, fig.height=6}
plot_bland_altman(blum_res$comparison$table, group_var="sample_id")
```

### Lechner
```{r lechner-ba, fig.height=6}
plot_bland_altman(lechner_res$comparison$table, group_var="sample_id")
```

### Luoma
```{r luoma-ba, fig.height=6}
plot_bland_altman(luoma_res$comparison$table, group_var="sample_id")
```

---

## 9  Failure Mode Analysis {.tabset}

For each sample, how many clonotypes share the same TRBV gene? Red-filled bars = TRBV families where a single clone accounts for >50% of cells — this is the main failure mode of the proxy.

### Blum
```{r blum-fail, fig.height=8, fig.width=14}
plot_within_trbv_diversity(blum_res$benchmark, group_var="sample_id")
```

### Lechner
```{r lechner-fail, fig.height=8, fig.width=14}
plot_within_trbv_diversity(lechner_res$benchmark, group_var="sample_id")
```

### Luoma
```{r luoma-fail, fig.height=8, fig.width=14}
plot_within_trbv_diversity(luoma_res$benchmark, group_var="sample_id")
```

---

## 10  Cross-Dataset Correlation Summary

```{r corr-summary}
corr_summary <- map_dfr(results_list, function(res) {
  comp <- res$comparison
  if (is.null(comp$spearman) || all(is.na(comp$spearman)))
    return(tibble(H_spearman_rho=NA, H_spearman_p=NA,
                  H_pearson_r=NA,   H_pearson_p=NA,
                  Clonality_spearman_rho=NA, n_samples=nrow(comp$table)))
  tibble(
    H_spearman_rho        = round(comp$spearman$estimate,           3),
    H_spearman_p          = round(comp$spearman$p.value,            4),
    H_pearson_r           = round(comp$pearson$estimate,            3),
    H_pearson_p           = round(comp$pearson$p.value,             4),
    Clonality_spearman_rho = round(comp$clonality_spearman$estimate, 3),
    n_samples             = nrow(comp$table)
  )
}, .id="Dataset")

corr_summary %>%
  kable(col.names = c("Dataset", "ρ (H)", "p (H)",
                       "r (H)", "p (H, Pearson)",
                       "ρ (Clonality)", "N samples"),
        caption = "Spearman ρ and Pearson r: TRBV proxy vs true clonotype diversity") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width=FALSE) %>%
  column_spec(2, bold=TRUE,
              color = ifelse(corr_summary$H_spearman_p < 0.05, "white", "black"),
              background = ifelse(corr_summary$H_spearman_p < 0.05, "#2166AC", "#CCCCCC"))
```

```{r corr-bar, fig.height=4, fig.width=7}
corr_summary %>%
  filter(!is.na(H_spearman_rho)) %>%
  mutate(Dataset = factor(Dataset, levels=c("blum","lechner","luoma")),
         sig = H_spearman_p < 0.05) %>%
  ggplot(aes(x=H_spearman_rho, y=Dataset, fill=Dataset)) +
    geom_col(width=0.55, alpha=0.9) +
    geom_errorbarh(aes(xmin=0, xmax=H_spearman_rho), height=0) +
    geom_text(aes(label=sprintf("ρ = %.2f%s", H_spearman_rho,
                                 ifelse(sig,"*",""))),
              hjust=-0.1, size=4) +
    geom_vline(xintercept=0, color="grey40") +
    scale_fill_manual(values=DATASET_COLORS, guide="none") +
    scale_x_continuous(limits=c(0, 1.15)) +
    labs(title   = "H_TRBV vs H_clonotype: Spearman ρ by Dataset",
         subtitle = "* p < 0.05",
         x = "Spearman ρ", y = NULL) +
    theme_bw(base_size=13)
```

---

## 11  Downsampled Sensitivity Analysis {.tabset}

Each dataset was re-analysed after downsampling to **`r DOWNSAMPLE_N` cells per sample**. Points should track the 1:1 line if estimates are stable across sampling depth.

```{r ds-scatter-fn}
ds_scatter <- function(res, dataset_id) {
  if (is.null(res$downsampled)) {
    return(ggplot() + annotate("text", x=0.5, y=0.5,
           label="Downsampling disabled (DOWNSAMPLE_N = 0)") + theme_void())
  }
  full_tbl <- build_summary_table(res$comparison) %>%
    select(sample_id, H_TRBV, H_clonotype) %>%
    rename(H_TRBV_full=H_TRBV, H_clono_full=H_clonotype)
  ds_tbl <- build_summary_table(res$downsampled$comparison) %>%
    select(sample_id, H_TRBV, H_clonotype) %>%
    rename(H_TRBV_ds=H_TRBV, H_clono_ds=H_clonotype)
  joined <- inner_join(full_tbl, ds_tbl, by="sample_id")

  p1 <- ggplot(joined, aes(H_TRBV_full, H_TRBV_ds, label=sample_id)) +
    geom_point(color=DATASET_COLORS[dataset_id], size=3) +
    geom_text_repel(size=2.8) +
    geom_abline(linetype="dashed", color="grey50") +
    labs(title="H_TRBV: Full vs Downsampled",
         x="Full", y=paste0("Downsampled (n=",DOWNSAMPLE_N,")")) +
    theme_bw(base_size=11)

  p2 <- ggplot(joined, aes(H_clono_full, H_clono_ds, label=sample_id)) +
    geom_point(color=DATASET_COLORS[dataset_id], size=3) +
    geom_text_repel(size=2.8) +
    geom_abline(linetype="dashed", color="grey50") +
    labs(title="H_clonotype: Full vs Downsampled",
         x="Full", y=paste0("Downsampled (n=",DOWNSAMPLE_N,")")) +
    theme_bw(base_size=11)

  p1 + p2
}
```

### Blum
```{r blum-ds, fig.height=5}
ds_scatter(blum_res, "blum")
```

### Lechner
```{r lechner-ds, fig.height=5}
ds_scatter(lechner_res, "lechner")
```

### Luoma
```{r luoma-ds, fig.height=5}
ds_scatter(luoma_res, "luoma")
```

---

## 12  Interpretation

```{r interp-table}
tibble(
  Scenario = c(
    "High ρ (>0.7), points near 1:1 line",
    "Low ρ or points far from 1:1 line",
    "High H_TRBV, Low H_clonotype",
    "Low H_TRBV, High H_clonotype",
    "Red-filled bars in failure mode plot"
  ),
  Meaning = c(
    "TRBV proxy is a reasonable diversity surrogate in this context (e.g. sorted CD8 T cells with no strong V-gene bias)",
    "Proxy breaks down; full TCR sequencing (CDR3) required for reliable clonality estimation",
    "Multiple TRBV genes used, but each harbours a single dominant clone — proxy overestimates diversity",
    "Few TRBV genes used, but each harbours many distinct CDR3s — proxy underestimates diversity (germline V-gene bias without clonal expansion)",
    "Single clone accounts for >50% of cells within that TRBV family — direct failure of the proxy"
  )
) %>%
  kable() %>%
  kable_styling(bootstrap_options=c("striped","hover"), full_width=TRUE) %>%
  row_spec(1, background="#d4edda") %>%
  row_spec(2:5, background="#fff3cd")
```

> **Bottom line:** The TRBV usage diversity proxy measures V-gene breadth/evenness — analogous to the NanoString nCounter TCR Diversity score. It **cannot** distinguish CDR3-level clonal expansion from V-gene usage bias. Use matched TCR sequencing (CDR3 clonotypes) for reliable clonality estimation. The proxy may still be useful as a cost-effective screening tool in contexts where matched TCR-seq is unavailable, provided its failure modes are understood.

---

<details>
<summary>Session info</summary>
```{r session-info}
sessionInfo()
```
</details>
